---
title: "using drop when subsetting"
output: github_document
date: '2023-02-23'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We often need to extract a subset of data out of an object.  We either want the same kind of object but with less data, or we want the actual data values.  This process is called 'extraction' or 'subsetting' and R provides tools for this. Another name for this is 'indexing' although purists will probably differ on that.  Who cares about them? 

Array like objects ( things laid out in regular patterns like arrays, matrices, data.frames, rasters, ...) can be index using `[`.  See `?[` for help on base R indexing.  

We'll make a data.frame as our example, but it could be a tibble (which is a data.frame), matrix, or array.

```{r}
n = 4
x = data.frame(A = seq_len(n), B = seq_len(n) * 5, C = seq_len(n) - 10)
x
```

Subsetting by indexing has us specify the `[rows, columns]` we want.

So the 3rd and 4th rows.

```{r}
x[3:4,]
```

Or we could select the 3rd and 4th rows but just columns A and B

```{r}
x[3:4, c("A", "B")]
```

In each case we get a subset of the same object type.  How about just getting column C for the 3rd and 4th rows?

```{r}
x[3:4, "C"]
```

Whoopsie!  It's just a vector! That's not the same type of object as `x`.  What gives?

It's an issue many languages have to face - what to do when a user selects one column... did they intended to subset and retain the same object type? Or did the want the actual values *in* that column to do some analysis with?  How is a poor computer to know?  Some languages opt to always assume users a lazy and just wanted the data in the column to do stuff with.  Other languages assume the user doesn't want to change object type.  R is the former.  So you have to explicitly tell R to *not* degenerate the data type from something complicated to something simple.  To do that you have to use the `drop = TRUE|FALSE` argument to `[` function (yes, it is a funny looking function!)   Here `drop` means if the result only has one column, then drop the dimensionality of the result to something simpler.   

Compare...

```{r}
x[3:4, "C", drop = TRUE]
```

... to ...

```{r}
x[3:4, "C", drop = FALSE]
```

Viola!  We now retain the original object type for the subset.

So, when should you use the `drop` argument?  Well, anytime you are building a function that will subset by indexing, you should at least consider adding `drop`.  But, of course, it depends upon the context of what you function needs to do.